---
title: "HELPS-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HELPS-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HELPS)
library(ncdf4)
library(raster)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Obtain example data:
```{r Step0, warning=FALSE, eval = FALSE}
get_example_data()
```

## 1. Daily gridded inputs: 

### Step1.1: calculate daily gridded heat stress

This step employs `HeatStress()` function, which takes multiple arguments. `SECTOR` denotes which sector the labor is involved in. The current version of **HELPS** supports 9 sectors, consists of 8 crop sectors (4 major crops by 2 irrigation practices) and 1 non-crop sector.
`HS` denotes the heat stress function that takes atmospheric variables as input and outputs heat stress level. The current version of **HELPS** covers 3 `HS()`, `WBGT_ESI()`, `WBGT_shade()`, and `WBGT_sun()`.
`YEAR_INPUT` denotes a year of interest.
`...` denotes the file name(s) of atmospheric variable projections. Note that the number of files need to be equal to the number of argument of `HS()`.
Example below takes monthly *hurs* (relative humidity), *tas* (near-surface air temperature), and *ps* (surface air pressure) and *WBGT_ESI* to calculate heat stress levels.

```{r Step1.1, warning=FALSE}
wbgt.sun.day <- HeatStress(SECTOR = "MAIZ_I", HS = WBGT_sun, YEAR_INPUT = 2024,
                    "../HELPS_Test_Data/hurs_day_GFDL-ESM2M_rcp60_r1i1p1_EWEMBI_20210101-20301231.nc4",
                    "../HELPS_Test_Data/tas_day_GFDL-ESM2M_rcp60_r1i1p1_EWEMBI_20210101-20301231.nc4",
                    "../HELPS_Test_Data/ps_day_GFDL-ESM2M_rcp60_r1i1p1_EWEMBI_20210101-20301231.nc4")
wbgt.sun.day
```

### Step1.2: calculate daily gridded physical work capacity 

This step employs `PWC()` function, which takes multiple arguments. 
`WBGT` denotes the output of `HeatStress()` from last step. 
`LHR` denotes the labor-heat response function that takes heat stress level and workload intensity as input and outputs physical work capacity (PWC). `workload` varies across alternative industries, we suggest 'high' for construction or agricultural work, 'moderate' for average manufacturing industry work, and 'low' for office desk work and service industries. 

```{r Step1.2, warning=FALSE}
pwc.hothaps.day <- PWC(WBGT = wbgt.sun.day,  LHR = LHR_Hothaps, workload = "high")
pwc.hothaps.day
```

### Step1.3: aggregate daily value to monthly value 

This step employs `DAY2MON()` function, which takes the daily output (a raster rack with 365/366 layers) from `HeatStress()` or `PWC()` as input, and outputs a raster rack of 12 layers with monthly mean.

```{r Step1.3, warning=FALSE}
wbgt.sun.mon <- DAY2MON(input_rack = wbgt.sun.day)
rm(wbgt.sun.day)
pwc.hothaps.mon <- DAY2MON(input_rack = pwc.hothaps.day)
rm(pwc.hothaps.day)
wbgt.sun.mon
pwc.hothaps.mon

```

### Step1.4: aggregate monthly value to annual value 

This step employs `MON2ANN()` function, which takes a raster rack with 12 layers of monthly mean as input, and outputs a data frame with grid-level annual mean. In this step, make sure the `SECTOR` input in `MON2ANN()` is identical to the `SECTOR` input in `HeatStress()`

```{r Step1.4, warning=FALSE}
wbgt.sun.ann <- MON2ANN(input_rack = wbgt.sun.mon, SECTOR = "MAIZ_I")
pwc.hothaps.ann <- MON2ANN(input_rack = pwc.hothaps.mon, SECTOR = "MAIZ_I")

summary(wbgt.sun.ann)
summary(pwc.hothaps.ann)
```

## 2. Monthly gridded inputs: 

### Step2.1: calculate monthly gridded heat stress

Similar to *Step1.1*, this steps employs`HeatStress()` function. Example below takes monthly *hurs* (relative humidity), *tas* (near-surface air temperature), and *rsds* (surface downwelling shortwave radiation) and *WBGT_ESI* to calculate heat stress levels.


```{r Step2.1, warning=FALSE}
esi.mon <- HeatStress(SECTOR = "MAIZ_I", HS = WBGT_ESI, YEAR_INPUT = 2024,
                    "../HELPS_Test_Data/hurs_mon_basd_CanESM5_W5E5v2_GCAM_ref_2015-2100.nc",
                    "../HELPS_Test_Data/tas_mon_basd_CanESM5_W5E5v2_GCAM_ref_2015-2100.nc",
                    "../HELPS_Test_Data/rsds_mon_basd_CanESM5_W5E5v2_GCAM_ref_2015-2100.nc")
esi.mon
```

### Step2.2: calculate daily gridded physical work capacity 

Similar to *Step1.2*, this step employs `PWC()` function to translate heat stress to PWC. Here we demonstrate the uncertainty in PWC introduced by alternative choices of `LHR()`.

```{r Step2.2.1, warning=FALSE}
start_t = Sys.time()
pwc.mon.foster <- PWC(WBGT = esi.mon,  LHR = LHR_Foster, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.foster
```
```{r Step2.2.2, warning=FALSE}
start_t = Sys.time()
pwc.mon.hothaps <- PWC(WBGT = esi.mon,  LHR = LHR_Hothaps, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.hothaps
```
```{r Step2.2.3, warning=FALSE}
start_t = Sys.time()
pwc.mon.niosh <- PWC(WBGT = esi.mon,  LHR = LHR_NIOSH, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.niosh
```
```{r Step2.2.4, warning=FALSE}
start_t = Sys.time()
pwc.mon.iso <- PWC(WBGT = esi.mon,  LHR = LHR_ISO, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.iso
```
### Step2.3: aggregate monthly value to annual value 

Same as *Step1.4*, make sure the `SECTOR` input in `MON2ANN()` is identical to the `SECTOR` input in `HeatStress()`

```{r Step2.3, warning=FALSE}
pwc.hothaps.ann2 <- MON2ANN(input_rack = pwc.mon.hothaps, SECTOR = "MAIZ_I")
summary(pwc.hothaps.ann2)
```

