---
title: "HELPS-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HELPS-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HELPS)
library(ncdf4)
library(raster)
library(dplyr)
library(ggplot2)
library(knitr)
library(sf)
library(ggpattern)
```

## Obtain example data:
```{r Step0, warning=FALSE, eval = FALSE}
get_example_data()
```

## 1. Daily gridded inputs: 

### Step1.1: calculate daily gridded heat stress

This step employs `HeatStress()` function, which takes multiple arguments. `SECTOR` denotes which sector the labor is involved in. The current version of **HELPS** supports 93 sectors, consists of 92 crop sectors (46 SPAM crops by 2 irrigation practices) and 1 non-crop sector.
`HS` denotes the heat stress function that takes atmospheric variables as input and outputs heat stress level. The current version of **HELPS** covers 3 `HS()`, `WBGT_ESI()`, `WBGT_shade()`, and `WBGT_sun()`.
`YEAR_INPUT` denotes a year of interest.
`...` denotes the file name(s) of atmospheric variable projections. Note that the number of files need to be equal to the number of argument of `HS()`.
Example below takes monthly *hurs* (relative humidity), *tas* (near-surface air temperature), and *ps* (surface air pressure) and *WBGT_ESI* to calculate heat stress levels.

```{r Step1.1, warning=FALSE}
wbgt.sun.day <- HeatStress(TempRes = "day", SECTOR = "SUNF_R", HS = WBGT_sun, YEAR_INPUT = 2024,
                    "../HELPS_Example_Data/hurs_example_day.nc",
                    "../HELPS_Example_Data/tas_example_day.nc",
                    "../HELPS_Example_Data/ps_example_day.nc")
wbgt.sun.day
```


```{r Step1.1.1, warning=FALSE, fig.width=10, fig.height=6}
wbgt.day.plot <- wbgt.sun.day[[183]] %>% as.data.frame(xy = T)

format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
wbgt.day.plot %>%
  setNames(c("x", "y", "value")) %>%
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = value)) +
  scale_fill_gradient2(low = "blue", mid = "grey95", high = "red", midpoint = 25,
                       na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "WBGT",
       title = "Heat stress for labor in rain-fed sunflower sector: 2024-07-01") +

  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```


### Step1.2: calculate daily gridded physical work capacity 

This step employs `PWC()` function, which takes multiple arguments. 
`WBGT` denotes the output of `HeatStress()` from last step. 
`LHR` denotes the labor-heat response function that takes heat stress level and workload intensity as input and outputs physical work capacity (PWC). `workload` varies across alternative industries, we suggest 'high' for construction or agricultural work, 'moderate' for average manufacturing industry work, and 'low' for office desk work and service industries. 

```{r Step1.2, warning=FALSE}
pwc.hothaps.day <- PWC(WBGT = wbgt.sun.day,  LHR = LHR_Hothaps, workload = "high")
pwc.hothaps.day
```
```{r Step1.2.1, warning=FALSE, fig.width=10, fig.height=6}
pwc.day.plot <- pwc.hothaps.day[[183]] %>% as.data.frame(xy = T)

format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
pwc.day.plot %>%
  setNames(c("x", "y", "value")) %>%
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = 100*(1-value))) +
  scale_fill_gradient(low = "grey95", high = "red",
                      na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Heat-induced PWC loss for labor in rain-fed sunflower sector: 2024-07-01") +

  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```

### Step1.3: aggregate daily value to monthly value 

This step employs `DAY2MON()` function, which takes the daily output (a raster rack with 365/366 layers) from `HeatStress()` or `PWC()` as input, and outputs a raster rack of 12 layers with monthly mean.

```{r Step1.3, warning=FALSE}
wbgt.sun.mon <- DAY2MON(input_rack = wbgt.sun.day)
# rm(wbgt.sun.day)
pwc.hothaps.mon <- DAY2MON(input_rack = pwc.hothaps.day)
# rm(pwc.hothaps.day)
wbgt.sun.mon
pwc.hothaps.mon

```
```{r Step1.3.1, warning=FALSE, fig.width=10, fig.height=6}
pwc.mon.plot <- pwc.hothaps.mon[[7]] %>% as.data.frame(xy = T)

format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
pwc.mon.plot %>%
  setNames(c("x", "y", "value")) %>% 
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = 100*(1-value))) +
  scale_fill_gradient(low = "grey95", high = "red",
                      na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Monthly heat-induced PWC loss for labor in rain-fed sunflower sector: 2024-07") +
  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```

### Step1.4: aggregate monthly value to annual value 

This step employs `MON2ANN()` function, which takes a raster rack with 12 layers of monthly mean as input, and outputs a data frame with grid-level annual mean. In this step, make sure the `SECTOR` input in `MON2ANN()` is identical to the `SECTOR` input in `HeatStress()`


```{r Step1.4, warning=FALSE}
wbgt.sun.ann <- MON2ANN(input_rack = wbgt.sun.mon, SECTOR = "SUNF_R")
pwc.hothaps.ann <- MON2ANN(input_rack = pwc.hothaps.mon, SECTOR = "SUNF_R")

summary(wbgt.sun.ann)
summary(pwc.hothaps.ann)
```

```{r Step1.4.1, warning=FALSE, fig.width=10, fig.height=6}
format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
pwc.hothaps.ann %>%
  setNames(c("x", "y", "value")) %>% 
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = 100*(1-value))) +
  scale_fill_gradient(low = "grey95", high = "red",
                      na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Annual heat-induced PWC loss for labor in rain-fed sunflower sector: 2024") +
  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```

## 2. Monthly gridded inputs: 

### Step2.1: calculate monthly gridded heat stress

Similar to *Step1.1*, this steps employs`HeatStress()` function. Example below takes monthly *hurs* (relative humidity), *tas* (near-surface air temperature), and *rsds* (surface downwelling shortwave radiation) and *WBGT_ESI* to calculate heat stress levels.

```{r Step2.1, warning=FALSE}
esi.mon <- HeatStress(TempRes = "month", SECTOR = "SUNF_R", HS = WBGT_ESI, YEAR_INPUT = 2024,
                    "../HELPS_Example_Data/hurs_example_month.nc",
                    "../HELPS_Example_Data/tas_example_month.nc",
                    "../HELPS_Example_Data/rsds_example_month.nc")
esi.mon
```

### Step2.2: calculate daily gridded physical work capacity 

Similar to *Step1.2*, this step employs `PWC()` function to translate heat stress to PWC. Here we demonstrate the uncertainty in PWC introduced by alternative choices of `LHR()`.

```{r Step2.2.1, warning=FALSE}
start_t = Sys.time()
pwc.mon.foster <- PWC(WBGT = esi.mon,  LHR = LHR_Foster, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.foster
```

```{r Step2.2.2, warning=FALSE}
start_t = Sys.time()
pwc.mon.hothaps <- PWC(WBGT = esi.mon,  LHR = LHR_Hothaps, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.hothaps
```
```{r Step2.2.3, warning=FALSE}
start_t = Sys.time()
pwc.mon.niosh <- PWC(WBGT = esi.mon,  LHR = LHR_NIOSH, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.niosh
```

```{r Step2.2.4, warning=FALSE}
start_t = Sys.time()
pwc.mon.iso <- PWC(WBGT = esi.mon,  LHR = LHR_ISO, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.iso
```
### Step2.3: aggregate monthly value to annual value 

Same as *Step1.4*, make sure the `SECTOR` input in `MON2ANN()` is identical to the `SECTOR` input in `HeatStress()`.

```{r Step2.3, warning=FALSE}
pwc.hothaps.ann2 <- MON2ANN(input_rack = pwc.mon.hothaps, SECTOR = "SUNF_R")
summary(pwc.hothaps.ann2)
```

## 3. Spatial aggregation: grid to region

```{r, warning=FALSE}
# aggregate grid level PWC to country level
ctry_pwc <- G2R(grid_annual_value = pwc.hothaps.ann, SECTOR = "SUNF_R", rast_boundary = country_raster)
# aggregate grid level PWC to GCAM water basin level
glu_pwc <- G2R(grid_annual_value = pwc.hothaps.ann, SECTOR = "SUNF_R", rast_boundary = reg_WB_raster)

```


### Step3.1: aggregate grid-level values to country-level values
```{r Step3.1, warning=FALSE, fig.width=10, fig.height=6}

GCAM_country %>% left_join(ctry_pwc, by = "region_id") -> df.plot

df.plot %>%  ggplot() +
  geom_sf_pattern(aes(fill = 100 * (1-value)),
                  pattern = ifelse(is.na(df.plot$value), "stripe", "none"),
                  pattern_color = "black",
                  pattern_density = 0.3,
                  pattern_spacing = 0.02) +
  scale_fill_gradient(low = "grey95", high = "red")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Country-level heat-induced PWC loss for labor in rain-fed sunflower sector: 2024") +
  theme_bw()
```

### Step3.2: aggregate grid-level values to GCAM waterbasin level values
```{r Step3.2, warning=FALSE, fig.width=10, fig.height=6}
reg_WB %>% left_join(glu_pwc, by = "region_id") -> df.plot

df.plot %>%  
  ggplot() +
  geom_sf_pattern(aes(fill = 100 * (1-value)),
                  pattern = ifelse(is.na(df.plot$value), "stripe", "none"),
                  pattern_color = "black",
                  pattern_density = 0.3,
                  pattern_spacing = 0.02) +
  scale_fill_gradient(low = "grey95", high = "red")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Water basin-level heat-induced PWC loss for labor in rain-fed sunflower sector: 2024") +
  theme_bw()
```
