---
title: "HELPS-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HELPS-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HELPS)
library(raster)
library(dplyr)
library(ggplot2)
library(knitr)
library(ggpattern)
```

## Gather input data:

### Download example data
Users can download a set of example data into folder `HELPS_Example_Data` by running `get_example_data()`. The example data includes grid-level daily and monthly near-surface air temperature (tas), daily and monthly near-surface relative humidity (hurs), daily surface air pressure (ps), and monthly grid-level surface downwelling shortwave radiation (rsds). *User Notice* in the *README* provides more details of input data `HELPS` is compatible with. 

```{r Step0, warning=FALSE, eval = FALSE}
get_example_data()
```

### Prepare your own input data

The `HELPS` package operates on 0.5 degree resolution, bias-corrected outputs from Earth System Models and General Circulation Models participating in the CMIP process. Several options to access such data exist:
- The [ISIMIP2b](https://data.isimip.org/search/tree/ISIMIP2b/InputData/climate/atmosphere/) repository contains outputs from CMIP5-era models (GFDL-ESM2M, HadGEM2-ES, IPSL-CM5A-LR, and MIROC5) for specific scenarios
- The [ISIMIP3b](https://data.isimip.org/search/tree/ISIMIP3b/InputData/climate/atmosphere/) repository contains outputs from CMIP6-era models (GFDL-ESM4, IPSL-CM6A-LR, MPI-ESM1-2-HR, MRI-ESM2-0, and UKSEM1-0-LL) for specific scenarios
- If a user wishes to explore other CMIP models or bias-correct model data against different observational data than that used by ISIMIP, the [BASD](https://github.com/JGCRI/basd) python package is available for ease of use. BASD implements an extension of the bias adjustment and statistical downscaling method used in ISIMIP3b ([ISIMIP3BASD](https://zenodo.org/records/4686991 and  https://gmd.copernicus.org/articles/12/3055/2019/), Lange 2021). The ISIMIP3BASD code base can also be used directly. For additional information, please see [here](https://www.isimip.org/documents/413/ISIMIP3b_bias_adjustment_fact_sheet_Gnsz7CO.pdf). 
- If a user wishes to explore novel scenarios not covered by CMIP models or the ISIMIP collection, we suggest a combination of [STITCHES](https://github.com/JGCRI/stitches) emulation and BASD bias correction and downscaling. 

HELPS has been tested with both ISIMIP data and outputs from BASD.

Besides the climate projections, `HELPS` also takes GGCMI Phase 3 crop calendar (Jägermeyr et al. 2021) (https://zenodo.org/records/5062513)  and SPAM harvested area (IFPRI, 2024) (https://doi.org/10.7910/DVN/SWPENT) data as input for temporal and spatial aggregation in the package.

`HELPS` provides measurement of crop specific labor heat stress exposure and heat-induced physical work capacity loss for 93 sectors, consisting of 46 SPAM crops by 2 irrigation practices (rain-fed and irrigated) and 1 noncrop sector. Below is the fUll list of sectors supported in `HELPS`, and refer to Table 1 for crop mapping: 
```
SECTOR_ALL <- c("WHEA_I", "RICE_I", "MAIZ_I", "SOYB_I", "BARL_I", "MILL_I", "PMIL_I", "SORG_I", "OCER_I",
                "POTA_I", "SWPO_I", "YAMS_I", "CASS_I", "BEAN_I", "CHIC_I", "COWP_I", "PIGE_I", "LENT_I",
                "GROU_I", "SUNF_I", "RAPE_I", "SESA_I", "SUGC_I", "SUGB_I", "COTT_I", "OFIB_I", "BANA_I",
                "PLNT_I", "CITR_I", "TROF_I", "TEMF_I", "TOMA_I", "ONIO_I", "VEGE_I", "ORTS_I", "OPUL_I",
                "CNUT_I", "OILP_I", "OOIL_I", "COFF_I", "RCOF_I", "COCO_I", "RUBB_I", "TEAS_I", "TOBA_I",
                "REST_I",
                "WHEA_R", "RICE_R", "MAIZ_R", "SOYB_R", "BARL_R", "MILL_R", "PMIL_R", "SORG_R", "OCER_R",
                "POTA_R", "SWPO_R", "YAMS_R", "CASS_R", "BEAN_R", "CHIC_R", "COWP_R", "PIGE_R", "LENT_R",
                "GROU_R", "SUNF_R", "RAPE_R", "SESA_R", "SUGC_R", "SUGB_R", "COTT_R", "OFIB_R", "BANA_R",
                "PLNT_R", "CITR_R", "TROF_R", "TEMF_R", "TOMA_R", "ONIO_R", "VEGE_R", "ORTS_R", "OPUL_R",
                "CNUT_R", "OILP_R", "OOIL_R", "COFF_R", "RCOF_R", "COCO_R", "RUBB_R", "TEAS_R", "TOBA_R",
                "REST_R",
                "NONCROP")
```
### Table 1. Crop mapping between data sources.
| SPAM | SPAM Crop Name    | ggcmi | ggcmi Crop Name|
|------|-------------------|-------|----------------|
| WHEA | Wheat             | wh    | wheat          |
| RICE | Rice              | ri    | rice           |
| MAIZ | Maize             | mai   | maize          |
| SOYB | Soybean           | soy   | soybean        |
| BARL | Barley            | bar   | barley         |
| MILL | Small Millet      | mil   | millet         |
| PMIL | Pearl Millet      | mil   | millet         |
| SORG | Sorghum           | sor   | sorghum        |
| OCER | Other Cereals     | rye   | rye            |
| POTA | Potato            | pot   | potato         |
| SWPO | Sweet Potato      | cas   | cassava        |
| YAMS | Yams              | cas   | cassava        |
| CASS | Cassava           | cas   | cassava        |
| BEAN | Bean              | bea   | beans          |
| CHIC | Chickpea          | pea   | field peas     |
| COWP | Cowpea            | pea   | beans          |
| PIGE | Pigeon Pea        | pea   | beans          |
| LENT | Lentil            | pea   | field peas     |
| GROU | Groundnut         | nut   | groundnut      |
| SUNF | Sunflower         | sun   | sunflower      |
| RAPE | Rapeseed          | rap   | rapeseed       |
| SESA | Sesame Seed       | sun   | sunflower      |
| SUGC | Sugarcane         | sgc   | sugar cane     |
| SUGB | Sugarbeet         | sgb   | sugar beet     |
| COTT | Cotton            | cot   | cotton         |
| OFIB | Other Fiber Crops |       |                |
| BANA | Banana            |       |                |
| PLNT | Plantain          |       |                |
| CITR | Citrus            |       |                |
| TROF | Other Tropical Fruit |    |                |
| TEMF | Temperate Fruit   |       |                |
| TOMA | Tomato            |       |                |
| ONIO | Onion             |       |                |
| VEGE | Other Vegetables  |       |                |
| ORTS | Other Roots       |       |                |
| OPUL | Other Pulses      |       |                |
| CNUT | Coconut           |       |                |
| OILP | Oilpalm           |       |                |
| OOIL | Other Oil Crops   |       |                |
| COFF | Arabic Coffee     |       |                |
| RCOF | Robust Coffee     |       |                |
| COCO | Cocoa             |       |                |
| RUBB | Rubber            |       |                |
| TEAS | Tea               |       |                |
| TOBA | Tobacco           |       |                |
| REST | Rest Of Crops     |       |                |



## 1. Daily gridded inputs: 

### Step1.1: calculate daily gridded heat stress

This step employs `HeatStress()` function, which takes multiple arguments. `SECTOR` denotes which sector the labor is involved in. The current version of **HELPS** supports 93 sectors, consists of 92 crop sectors (46 SPAM crops by 2 irrigation practices) and 1 non-crop sector.
`HS` denotes the heat stress function that takes atmospheric variables as input and outputs heat stress level. The current version of **HELPS** covers 3 `HS()`, `WBGT_ESI()`, `WBGT_shade()`, and `WBGT_sun()`. 
`WBGT_ESI`, Environmental Stress Index, takes hurs, tas, and rsds as inputs to approximate WBGT.
`WBGT_shade`, from Dunne et al. (2013), takes hurs, tas, and ps as inputs to approximate WBGT. This method assumes full shade and night adaptation and optimization of structures, clothing, activity scheduling.
`WBGT_sun`, is `WBGT_shade + 3`, that approximate the WBGT without shade.

`YEAR_INPUT` denotes a year of interest.
`...` denotes the file name(s) of atmospheric variable projections. Note that the number of files need to be equal to the number of argument of `HS()`.
Example below takes monthly *hurs* (relative humidity), *tas* (near-surface air temperature), and *ps* (surface air pressure) and *WBGT_ESI* to calculate heat stress levels.

```{r Step1.1, warning=FALSE}
wbgt.sun.day <- HeatStress(TempRes = "day", SECTOR = "SUNF_R", HS = WBGT_sun, YEAR_INPUT = 2024,
                    "../HELPS_Example_Data/hurs_example_day.nc",
                    "../HELPS_Example_Data/tas_example_day.nc",
                    "../HELPS_Example_Data/ps_example_day.nc")
wbgt.sun.day
```


```{r Step1.1.1, warning=FALSE, fig.width=10, fig.height=6}
wbgt.day.plot <- wbgt.sun.day[[183]] %>% as.data.frame(xy = T)

format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
wbgt.day.plot %>%
  setNames(c("x", "y", "value")) %>%
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = value)) +
  scale_fill_gradient2(low = "blue", mid = "grey95", high = "red", midpoint = 25,
                       na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "WBGT",
       title = "Heat stress for labor in rain-fed sunflower sector: 2024-07-01") +

  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```


### Step1.2: calculate daily gridded physical work capacity 

This step employs `PWC()` function, which takes multiple arguments. 
`WBGT` denotes the output of `HeatStress()` from last step. 
`LHR` denotes the labor-heat response function that takes heat stress level and workload intensity as input and outputs physical work capacity (PWC). `workload` varies across alternative industries, we suggest 'high' for construction or agricultural work, 'moderate' for average manufacturing industry work, and 'low' for office desk work and service industries. 
There are 4 built-in `LHR` functions in `HELPS`. 
`LHR_Hothaps` is based on the *high occupational temperature health and productivity suppression (Hothaps)* programme, derived by Kjellstrom et al. (2018).
`LHR_NOISH` calculate the labor productivity based on the National Institute for Occupational Safety and Health (NIOSH) standard.
`LHR_ISO` calculate the labor productivity based on the ISO7243 standard.
`LHR_Foster` calculate the labor productivity based on Foster et al (2021) Table 3. 
Note that `LHR_Foster` does not need workload as input, while the rest three LHRs need.


```{r Step1.2, warning=FALSE}
pwc.hothaps.day <- PWC(WBGT = wbgt.sun.day,  LHR = LHR_Hothaps, workload = "high")
pwc.hothaps.day
```
```{r Step1.2.1, warning=FALSE, fig.width=10, fig.height=6}
pwc.day.plot <- pwc.hothaps.day[[183]] %>% as.data.frame(xy = T)

format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
pwc.day.plot %>%
  setNames(c("x", "y", "value")) %>%
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = 100*(1-value))) +
  scale_fill_gradient(low = "grey95", high = "red",
                      na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Heat-induced PWC loss for labor in rain-fed sunflower sector: 2024-07-01") +

  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```

### Step1.3: aggregate daily value to monthly value 

This step employs `DAY2MON()` function, which takes the daily output (a raster rack with 365/366 layers) from `HeatStress()` or `PWC()` as input, and outputs a raster rack of 12 layers with monthly mean.

```{r Step1.3, warning=FALSE}
wbgt.sun.mon <- DAY2MON(input_rack = wbgt.sun.day)
# rm(wbgt.sun.day)
pwc.hothaps.mon <- DAY2MON(input_rack = pwc.hothaps.day)
# rm(pwc.hothaps.day)
wbgt.sun.mon
pwc.hothaps.mon

```
```{r Step1.3.1, warning=FALSE, fig.width=10, fig.height=6}
pwc.mon.plot <- pwc.hothaps.mon[[7]] %>% as.data.frame(xy = T)

format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
pwc.mon.plot %>%
  setNames(c("x", "y", "value")) %>% 
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = 100*(1-value))) +
  scale_fill_gradient(low = "grey95", high = "red",
                      na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Monthly heat-induced PWC loss for labor in rain-fed sunflower sector: 2024-07") +
  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```

### Step1.4: aggregate monthly value to annual value 

This step employs `MON2ANN()` function, which takes a raster rack with 12 layers of monthly mean as input, and outputs a data frame with grid-level annual mean. In this step, make sure the `SECTOR` input in `MON2ANN()` is identical to the `SECTOR` input in `HeatStress()`.


```{r Step1.4, warning=FALSE}
wbgt.sun.ann <- MON2ANN(input_rack = wbgt.sun.mon, SECTOR = "SUNF_R")
pwc.hothaps.ann <- MON2ANN(input_rack = pwc.hothaps.mon, SECTOR = "SUNF_R")

summary(wbgt.sun.ann)
summary(pwc.hothaps.ann)
```

```{r Step1.4.1, warning=FALSE, fig.width=10, fig.height=6}
format_longitude <- function(lon) {
  sapply(lon, function(x) {
    if (x > 0) paste0(x, "°E")
    else if (x < 0) paste0(abs(x), "°W")
    else paste0(x, "°") # Handle 0° specifically
  })
}

format_latitude <- function(lat) {
  sapply(lat, function(x) {
    if (x > 0) paste0(x, "°N")
    else if (x < 0) paste0(abs(x), "°S")
    else paste0(x, "°") # Handle 0° specifically
  })
}
pwc.hothaps.ann %>%
  setNames(c("x", "y", "value")) %>% 
  ggplot() +
  geom_tile(aes(x = x, y = y, fill = 100*(1-value))) +
  scale_fill_gradient(low = "grey95", high = "red",
                      na.value = "white")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Annual heat-induced PWC loss for labor in rain-fed sunflower sector: 2024") +
  scale_x_continuous(
    breaks = seq(-180, 180, by = 60), # Longitude breaks
    labels = format_longitude         # Apply custom labels
  ) +
  scale_y_continuous(
    breaks = seq(-90, 90, by = 30),    # Latitude breaks
    labels = format_latitude           # Apply custom latitude labels
  ) +
  theme_bw()
```

## 2. Monthly gridded inputs: 

### Step2.1: calculate monthly gridded heat stress

Similar to *Step1.1*, this steps employs`HeatStress()` function. Example below takes monthly *hurs* (relative humidity), *tas* (near-surface air temperature), and *rsds* (surface downwelling shortwave radiation) and *WBGT_ESI* to calculate heat stress levels.

```{r Step2.1, warning=FALSE}
esi.mon <- HeatStress(TempRes = "month", SECTOR = "SUNF_R", HS = WBGT_ESI, YEAR_INPUT = 2024,
                    "../HELPS_Example_Data/hurs_example_month.nc",
                    "../HELPS_Example_Data/tas_example_month.nc",
                    "../HELPS_Example_Data/rsds_example_month.nc")
esi.mon
```

### Step2.2: calculate daily gridded physical work capacity 

Similar to *Step1.2*, this step employs `PWC()` function to translate heat stress to PWC. Here we demonstrate the uncertainty in PWC introduced by alternative choices of `LHR()`.

```{r Step2.2.1, warning=FALSE}
start_t = Sys.time()
pwc.mon.foster <- PWC(WBGT = esi.mon,  LHR = LHR_Foster, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.foster
```

```{r Step2.2.2, warning=FALSE}
start_t = Sys.time()
pwc.mon.hothaps <- PWC(WBGT = esi.mon,  LHR = LHR_Hothaps, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.hothaps
```
```{r Step2.2.3, warning=FALSE}
start_t = Sys.time()
pwc.mon.niosh <- PWC(WBGT = esi.mon,  LHR = LHR_NIOSH, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.niosh
```

```{r Step2.2.4, warning=FALSE}
start_t = Sys.time()
pwc.mon.iso <- PWC(WBGT = esi.mon,  LHR = LHR_ISO, workload = "high")
end_t = Sys.time()
end_t - start_t
pwc.mon.iso
```
### Step2.3: aggregate monthly value to annual value 

Same as *Step1.4*, make sure the `SECTOR` input in `MON2ANN()` is identical to the `SECTOR` input in `HeatStress()`.

```{r Step2.3, warning=FALSE}
pwc.hothaps.ann2 <- MON2ANN(input_rack = pwc.mon.hothaps, SECTOR = "SUNF_R")
summary(pwc.hothaps.ann2)
```

## 3. Spatial aggregation: grid to region

```{r, warning=FALSE}
# aggregate grid level PWC to country level
ctry_pwc <- G2R(grid_annual_value = pwc.hothaps.ann, SECTOR = "SUNF_R", rast_boundary = country_raster)
# aggregate grid level PWC to GCAM water basin level
glu_pwc <- G2R(grid_annual_value = pwc.hothaps.ann, SECTOR = "SUNF_R", rast_boundary = reg_WB_raster)

```


### Step3.1: aggregate grid-level values to country-level values
```{r Step3.1, warning=FALSE, fig.width=10, fig.height=6}

GCAM_country %>% left_join(ctry_pwc, by = "region_id") -> df.plot

df.plot %>%  ggplot() +
  geom_sf_pattern(aes(fill = 100 * (1-value)),
                  pattern = ifelse(is.na(df.plot$value), "stripe", "none"),
                  pattern_color = "black",
                  pattern_density = 0.3,
                  pattern_spacing = 0.02) +
  scale_fill_gradient(low = "grey95", high = "red")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Country-level heat-induced PWC loss for labor in rain-fed sunflower sector: 2024") +
  theme_bw()
```

### Step3.2: aggregate grid-level values to GCAM waterbasin level values
```{r Step3.2, warning=FALSE, fig.width=10, fig.height=6}
reg_WB %>% left_join(glu_pwc, by = "region_id") -> df.plot

df.plot %>%  
  ggplot() +
  geom_sf_pattern(aes(fill = 100 * (1-value)),
                  pattern = ifelse(is.na(df.plot$value), "stripe", "none"),
                  pattern_color = "black",
                  pattern_density = 0.3,
                  pattern_spacing = 0.02) +
  scale_fill_gradient(low = "grey95", high = "red")+
  labs(x = "longitude", y = "latitude", fill = "PWC \nloss(%)",
       title = "Water basin-level heat-induced PWC loss for labor in rain-fed sunflower sector: 2024") +
  theme_bw()
```
